---
title: "About Class"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
```

**Office Location**: CBLS 377

**Telephone**: 4-5404

**Email**: rsschwartz@uri.edu  
Please note that I will do my best to respond promptly between 9-5 on weekdays.

**Student Hours**: Make an appointment for any available slot on my   [calendar](https://tinyurl.com/schwartz-calendar)
To talk with me during other times please set up a meeting by email.

**Class Days / Time**: M 3-4pm Beaupre 215

**Prerequisites**: BIO 262 (previous or concurrent)

**Required text**:
Whitlock and Schluter. 2015. The analysis of biological data. 2nd or 3rd ed.

**Other equipment / material requirements**:
Laptop, Tablet, or Chromebook with internet access. Please contact the instructor if you do not have a laptop and purchasing one would be a financial difficulty. Laptops are available for use in class upon request.

**Exams**:
The course will be divided into four sections.
For each section, points will be earned based on a short report 
based on results from in-class work. 
These "exams" will build on skills and concepts learned in previous weeks.
The exams will be worth 150, 250, 300, and 300 points, respectively.
The final exam will be assigned the last day of class and due at the end of the finals period.
All exams are take-home, open book, and open note.
You may consult with the professor and your fellow students but your 
work should be produced independently, including code and captions.

**Grades**:

```{r}
Grade <- c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "F") 
Points <- c("940-1000", "900-939", "870-899", "830-869", "800-829", "770-799", "730-769", 
            "700-729", "670-699", "600-669", "<600")

kable(cbind(Grade,Points), "html") %>%  
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))

```
